\bf The usage of bootstrap method in shape-restricted regression

### 2023-08-24 Manuscript/manuscript.tex
Add files via upload

Bootstrap is a powerful and important resampling method for estimating population parameters or assessing the accuracy of a statistical procedure by repeatedly sampling the data with replacement. Its effectiveness have been widely recognized in linear regression analysis \cite{efron1979bootstrap}, and non-linear regression analysis \cite{davidson1999bootstrap}. By bootstrap method, we can estimate the standard errors of the regression coefficients. Even for unknown underlying population distributions, bootstrap method can resample the data with replacement and computes the regression coefficients for each resampled dataset. \cite{hall2013simple}By repeating this process many times, it provides a valid empirical estimate of the standard errors in both linear cases \cite{efron1985bootstrap} and nonlinear cases\cite{wong2019bootstrap}, thus it is valuable when the assumptions of traditional standard error estimation techniques are violated. It is also a valid method used to construct confidence intervals for regression coefficients. \cite{efron1985bootstrap}By resampling the dataset, bootstrap can determining the percentile intervals of these coefficients, and also can obtain approximate confidence intervals for the regression parameters. \cite{cui2012evaluating} Bootstrap can also construct the approximate confidence interval for certain time points f(t)s in the function by simulating f'(t) certain times in certain time points. \cite{horowitz2018bootstrap} However, when dealing with shape-restricted regressions, where the coefficients are forced to adhere to specific constraints such as monotonicity or curvature, the effectiveness of bootstrap is not proved to be guaranteed. In this paper, we will demonstrate the usage of the bootstrap method in shape-restricted regression and proof the effectiveness of bootstrap for constructing pointwise confidence intervals in shape-restricted regression empirically.

@@

\section{Bootstrap Method for Constructing Confidence Intervals}
\label{Bootstrap Method for Constructing Confidence Intervals}

The bootstrap method is a powerful resampling technique used to construct confidence intervals for population parameters or estimates when the underlying distribution is unknown or difficult to determine. \cite{hall2013simple}It is useful in regression analysis to estimate the uncertainty associated with the regression coefficients or for certain time points inside the regression function. Specifically, the bootstrap method basically contains three steps:

\begin{enumerate}[label=\arabic*.]
    \item Let \(v\) represent a population parameter of interest (e.g., mean, median, standard deviation, etc.), which is drawn from an unknown population distribution \(F\).
    \item Generate a bootstrap sample \(x_1, x_2, \ldots, x_n\) from the population, and they are referred to as the original sample dataset.
    \item Let \(u\) represent a statistic calculated from the bootstrap original sample dataset.
    \item Using the data in the original sample dataset as the "population," perform nonparametric resampling with replacement to obtain a resampled sample (also known as the Bootstrap sample), denoted as \(x_1^*, x_2^*, \ldots, x_n^*\) (the number of data points in the Bootstrap sample must be the same as the number of data points in the original sample).
    \item Let \(u^*\) represent the statistic computed by the Bootstrap sample datasets, and this will be the bootstrap estimated statistic for the population statistic.

@@

Bootstrap method has been proved that it is an excellent method to estimate statistic \cite{efron1979bootstrap}, the \(u^*\) we obtained from the resampling dataset is a good estimator for the \(v\) as a population parameter. According to the bootstrap method, we can also calculate the confidence interval for the parameters, or the coefficients of a certain regression, or for certain time points in the regression. In this article, we will focus on the method of construct a pointwise confidence interval by bootstrap. To construct a pointwise confidence interval using the bootstrap method, the procedure involves the following steps:

\begin{enumerate}[label=\arabic*.]
    \item First we observed a specific time point \(t\) that we are interested and its corresponding true value \(f(t)\), and build them as a pair \(t, f(t)\).
    \item According to the dataset, fit \(\hat{f}\) as the estimated regression for the dataset.
    \item Generate a bootstrap sample \(x_1, x_2, \ldots, x_n\) from the population by sampling with replacement, and they are referred to as the original sample dataset. 
    \item Then  \( [\hat{f}_n(t) - \hat{q}_{1-\alpha/2,n-1/3}, \hat{f}_n(t) - \hat{q}_{\alpha/2,n-1/3} ]\)  would be the bootstrap estimated pointwise confidence interval where \(\hat{q}_\alpha\) denoted the \(\alpha\)th quantile of the time point \(t\).

It has been proved that the bootstrap method can be used as a good estimator for the pointwise confidence interval in linear regressions and non-linear regressions\cite{efron1979bootstrap}\cite{davidson1999bootstrap}, and some examples in certain configurations are given.\cite{ruhe2019bootstrap}\cite{ma2019inference}\cite{dugas2010pointwise} But in certain special cases(i.e. shape-restricted splines regressions) there is no certain resource have not prove that it is a good method to build pointwise confidence interval via bootstrap.

@@

We define three functions: \(y = 0.25(x - 0.9)\) , \(y = 0.02(x^3+1.2)\) , and \( y = 0.1 (5\Phi((x - 2) / 0.3) + 1)\) , where \(\Phi\) is the cumulative distribution function of the standard normal distribution. For those functions, the validity of the bootstrap method to build the confidence intervals for the coefficients and standard error has already been proved. \cite{jieying2022heteroscedastic} And in this article we want to further prove the validity of the bootstrap method to build the pointwise confidence intervals for certain time points in the domain of the function. First of all, we randomly generate 1000 points \(x\) in the domain of \([0,30]\), and we calculate its corresponding values \(y(x)\), and make simulation data points \(f'(x)\) by add noise to the true value \(f(x)\). The noise is added from a standard normal distribution with \(\mu = 0\) and \(\sigma = \) \(1\). And then we make the paired values \((x, y\text{ with noise})\) be the dataset, while we know the true values of \(f(x)\) as well.

\begin{figure}[H]
  \centering
  \includegraphics[width=0.8\textwidth]{DataAndRegressions.png}
  \caption{The generated dataset, true functions and fitted i-splines regressions: the shaded green curve represents the true functions, the pink points are the data pair \((x, f'(x))\) where \(f'(x)\) is generated by \(f(x)\) add noise, and the red curves are the fitted i-splines regression curve with degrees of freedom\(= 6,10,15\), from top to bottom}
  \label{fig:DataAndRegressions}
\end{figure}

We use quadratic i-splines function to enforce the monotonicity of the function and we picked degrees of freedom 6,10,15 to build the regression for the dataset. The generated dataset and regressions is shown in Figure ~\ref{fig:DataAndRegressions}. In each configuration, we select 1, 5, 10, 15, 20, 25, 30 as the time points from the dataset and perform bootstrap resampling 1000 times to estimate the \(95\%\) quantile pointwise confidence intervals for each of the selected time points. We repeat this process 1000 times to see how many times these confidence intervals can contain the true result. The results are shown in Table ~\ref{Table}.

\begin{table}[ht]
  \centering
  \label{Table}
  \caption{Results of functions for different degrees of freedom}
  \begin{tabular}{|c|c|c|c|c|}
    \hline
    \textbf{x} & \textbf{f(x)} & \textbf{CP of df\_6} & \textbf{CP of df\_10} & \textbf{CP of df\_15} \\
    \hline
    \multicolumn{5}{|c|}{\(y = 0.25(x - 0.9)\)} \\
    \hline
    1 & 0.025 & 0.936 & 0.944 & 0.939 \\
    \hline
    5 & 1.025 & 0.960 & 0.958 & 0.948 \\
    \hline
    10 & 2.275 & 0.956 & 0.950 & 0.951 \\
    \hline
    15 & 3.525 & 0.959 & 0.954 & 0.954 \\
    \hline
    20 & 4.775 & 0.941 & 0.949 & 0.961 \\
    \hline
    25 & 6.025 & 0.953 & 0.946 & 0.942 \\
    \hline
    30 & 7.275 & 0.919 & 0.955 & 0.947 \\
    \hline
    \multicolumn{5}{|c|}{\(y = 0.02(x^3+1.2)\)} \\
    \hline
    1 & 0.025 & 0.893 & 0.940 & 0.959 \\
    \hline
    5 & 1.025 & 0.932 & 0.948 & 0.958 \\
    \hline
    10 & 2.275 & 0.937 & 0.948 & 0.941 \\
    \hline
    15 & 3.525 & 0.942 & 0.951 & 0.953 \\
    \hline
    20 & 4.775 & 0.939 & 0.951 & 0.957 \\
    \hline
    25 & 6.025 & 0.952 & 0.959 & 0.963 \\
    \hline
    30 & 7.275 & 0.947 & 0.943 & 0.943 \\
    \hline
    \multicolumn{5}{|c|}{\(y = 0.1 (5\Phi((x - 2) / 0.3) + 1)\)} \\
    \hline
    1 & 0.025 & 0.686 & 0.847 & 0.901 \\
    \hline
    5 & 1.025 & 0.279 & 0.800 & 0.852 \\
    \hline
    10 & 2.275 & 0.947 & 0.937 & 0.942 \\
    \hline
    15 & 3.525 & 0.937 & 0.953 & 0.961 \\
    \hline
    20 & 4.775 & 0.929 & 0.950 & 0.957 \\
    \hline
    25 & 6.025 & 0.953 & 0.924 & 0.922 \\
    \hline
    30 & 7.275 & 0.919 & 0.858 & 0.832 \\
    \hline
  \end{tabular}
\end{table}

In table ~\ref{Table}, \(x\) represents the certain time point we selected in the domain of the function, and \(f(x)\) is its true value calculated by the certain known function. CP represents coverage percentage of pointwise confidence interval we built by the bootstrap methods under certain degrees of freedom. We can see that in all those functions and degrees of freedom we have given, the bootstrap method gives pointwise confidence intervals coverage that are closed to the nominal level \(95\%\).

>>>>

In certain area and cases, apply shape restrictions, such as monotonicity or convexity to estimate a certain function is reasonable and not uncommon in many fields.\cite{guo2019smooth} There are several methods to build shape-restricted regressions to do analysis the regressions, like quadratic programming.\cite{meyer2013simple} However, due to shape restrictions, it is not quite accurate to estimate the uncertainty of the parameters with normal methods like MLE (Maximum likelihood estimation) since parameters will sometimes lies on the boundary. In this paper, we consider to use bootstrap method to replace MLE method to estimate the uncertainty of the parameters, specifically, of certain time points in the regression curve. Although the bootstrap method is a valid method used to construct confidence intervals for regression coefficients, \cite{efron1985bootstrap}and its effectiveness have been widely recognized in linear regression analysis \cite{efron1979bootstrap}, and non-linear regression analysis \cite{davidson1999bootstrap}. However, when dealing with shape-restricted regressions, where the coefficients are forced to adhere to specific constraints such as monotonicity or curvature, the effectiveness of bootstrap is not proved to be guaranteed. In this paper, we will demonstrate the usage of the bootstrap method in shape-restricted regression and proof the effectiveness of bootstrap for constructing pointwise confidence intervals in shape-restricted regression empirically.

@@

\section{Bootstrap Method review}
\label{Bootstrap Method for Constructing Confidence Intervals}

The bootstrap is a powerful and important resampling method for estimating population parameters or assessing the accuracy of a statistical procedure by repeatedly sampling the data with replacement. Its effectiveness have been widely recognized in linear regression analysis \cite{efron1979bootstrap}, and non-linear regression analysis \cite{davidson1999bootstrap}. By bootstrap method, we can estimate the standard errors and confidence intervals for the regression coefficients. Even for unknown underlying population distributions, bootstrap method can resample the data with replacement and computes the regression coefficients for each resampled dataset. \cite{hall2013simple}By repeating this process many times, it provides a valid empirical estimate of the standard errors in both linear cases \cite{efron1985bootstrap} and nonlinear cases\cite{wong2019bootstrap}, thus it is valuable when the assumptions of traditional standard error estimation techniques are violated. It is also a valid method used to construct confidence intervals for regression coefficients. \cite{efron1985bootstrap}By resampling the dataset, bootstrap can determining the percentile intervals of these coefficients, and also can obtain approximate confidence intervals for the regression parameters. \cite{cui2012evaluating} Bootstrap can also construct the approximate confidence interval for certain time points f(t)s in the function by simulating f'(t) certain times in certain time points. \cite{horowitz2018bootstrap} Thus it is useful in regression analysis to estimate the uncertainty associated with the regression coefficients or for certain time points inside the regression function. \cite{Efron2011-sn}For convenience, in here we just talk about simple regression model which only contains one independent variable and one dependent variable. Specifically, the bootstrap method basically contains the following steps:

\begin{enumerate}[label=\arabic*.]
    \item Suppose we observed specific independent data points \(x_1, x_2, \ldots, x_n\) from a dataset, and for convenience we placed them inside a vector \(X = (x_1, x_2, \ldots, x_n)\), and since we want to construct a regression for the dataset, these data points are supposed to have a corresponding \(y\) values as dependent variable, and this makes the data points becomes pairs as \(X = (x_1,y_1) , (x_2,y_2),  \ldots, (x_n,y_n)\), and we call this as original dataset \(X\). Then we use the original dataset \(X)\) to create a simple regression model \(y = \beta_0 + \beta_1 x\).
    \item From the original dataset \(X\), generate \(n = B\) bootstrap samples, and placed them inside a dataset denoted as \(X^* = x_1^*, x_2^*, \ldots, x_n^*\), each bootstrap samples are obtained by sampling inside the original dataset \(X\) with replacement.
    \item With this bootstrap dataset, we can do the inference for any statistics we are interested in the original dataset. For exsample, if we are interested in the standard error \(s(x)\) for the original dataset, we may use the bootstrap dataset to inference it. The bootstrap estimate of standard error is the standard deviation calculated by the bootstrap dataset \(X^*\):
    \begin{equation}
      \hat{se_{boot}} = \{\sum_{b=1}^{B}[s(X^*)-s(*)]^2/(B-1)\}^{1/2}
    \end{equation}

    \item Using the data in the original sample dataset as the "population," perform nonparametric resampling with replacement to obtain a resampled sample (also known as the Bootstrap sample), denoted as \(x_1^*, x_2^*, \ldots, x_n^*\) (the number of data points in the Bootstrap sample must be the same as the number of data points in the original sample).
    \item If we are interested in confidence interval for certain regression coefficients or time points, we can also use bootstrap method to construct it, in this situation, \( [\hat{f}_n(t) - \hat{q}_{1-\alpha/2,n-1/3}, \hat{f}_n(t) - \hat{q}_{\alpha/2,n-1/3} ]\)  would be the bootstrap estimated pointwise confidence interval where \(\hat{q}_\alpha\) denoted the \(\alpha\)th quantile of the time point \(t\).

@@

Bootstrap method has been proved that it is an excellent method to estimate statistic \cite{efron1979bootstrap}, the \(u^*\) we obtained from the resampling dataset is a good estimator for the \(v\) as a population parameter. According to the bootstrap method, we can also calculate the confidence interval for the parameters, or the coefficients of a certain regression, or for certain time points in the regression. In this article, we will focus on the method of construct a pointwise confidence interval by bootstrap. It has been proved that the bootstrap method can be used as a good estimator for the pointwise confidence interval in linear regressions and non-linear regressions\cite{efron1979bootstrap}\cite{davidson1999bootstrap}, and some examples in certain configurations are given.\cite{ruhe2019bootstrap}\cite{ma2019inference}\cite{dugas2010pointwise} But in certain special cases(i.e. shape-restricted splines regressions) there is no certain resource have not prove that it is a good method to build pointwise confidence interval via bootstrap.

@@

We define three functions: \(y = 0.25(x - 0.9)\) , \(y = 0.02(x^3+1.2)\) , and \(y = 0.1(5\Phi((x - 2)/0.3) + 1)\) , where \(\Phi\) is the cumulative distribution function of the standard normal distribution. For those functions, the validity of the bootstrap method to build the confidence intervals for the coefficients and standard error has already been proved. \cite{jieying2022heteroscedastic} And in this article we want to further prove the validity of the bootstrap method to build the pointwise confidence intervals for certain time points in the domain of the function. First of all, we randomly generate some data points \(x\) in the domain of \([0,4]\), and we calculate its corresponding values \(y(x)\), and make simulation data points \(f'(x)\) by add noise to the true value \(f(x)\). The noise is added from a standard normal distribution with \(\mu = 0\) and \(\sigma = \) \(1\). And then we make the paired values \((x, y\text{ with noise})\) be the dataset, while we know the true values of \(f(x)\) as well.

\begin{figure}[H]
  \centering
  \includegraphics[width=0.8\textwidth]{DataPointsGeneration.png}
  \caption{The generated dataset while n=100, true functions and fitted i-splines regressions: the shaded green curve represents the true functions, the pink points are the data pair \((x, f'(x))\) where \(f'(x)\) is generated by \(f(x)\) add noise, and the red curves are the fitted i-splines regression curve with degrees of freedom\(= 6,10,15\), from top to bottom}
  \label{fig:DataAndRegressions}
\end{figure}

We use quadratic i-splines function to enforce the monotonicity of the function and we picked degrees of freedom 6 and 10 to build the regression for the dataset. The generated dataset and regressions is shown in Figure ~\ref{fig:DataAndRegressions}. While according to the law of large number, we also expect that the result will be closer to \(95\%\) while n gets larger, so we also did a simulation for \(n = 1000\) to see the numeric coverage percentiles. In each configuration, we select \(0.5,1,1.5,2,2.5,3,3.5,4\) as the time points from the dataset and perform bootstrap resampling 1000 times to estimate the \(95\%\) quantile pointwise confidence intervals for each of the selected time points. We repeat this process 1000 times to see how many times these confidence intervals can contain the true result. The results are shown in Table ~\ref{Table} and Table ~/ref{2}.

\begin{table}[ht]
  \centering
  \caption{Results for Different Degrees of Freedom (\(n = 100\))}
  \label{Table1}
  \begin{tabular}{|c|c|c|c|}
    \hline
    \textbf{x} & \textbf{Function} & \textbf{CP of df\_6} & \textbf{CP of df\_10} \\
    \hline
    \multicolumn{4}{|c|}{\(y = 0.25(x - 0.9)\)} \\
    \hline
    0.5 & -0.100 & 0.952 & 0.959 \\
    \hline
    1.0 & 0.025 & 0.956 & 0.948 \\
    \hline
    1.5 & 0.150 & 0.959 & 0.940\\
    \hline
    2.0 & 0.275 & 0.956 & 0.953 \\
    \hline
    2.5 & 0.400 & 0.952 & 0.957 \\
    \hline
    3.0 & 0.525 & 0.950 & 0.953 \\
    \hline
    3.5 & 0.650 & 0.945 & 0.950 \\
    \hline
    4.0 & 0.775 & 0.926 & 0.908 \\
    \hline
    \multicolumn{4}{|c|}{\(y = 0.02(x^3+1.2)\)} \\
    \hline
    0.5 & 0.027 & 0.938 & 0.937 \\
    \hline
    1.0 & 0.044 & 0.946 & 0.952 \\
    \hline
    1.5 & 0.092 & 0.932 & 0.943 \\
    \hline
    2.0 & 0.184 & 0.936 & 0.945 \\
    \hline
    2.5 & 0.337 & 0.942 & 0.948 \\
    \hline
    3.0 & 0.564 & 0.960 & 0.946 \\
    \hline
    3.5 & 0.882 & 0.940 & 0.943 \\
    \hline
    4.0 & 1.304 & 0.907 & 0.907 \\
    \hline
    \multicolumn{4}{|c|}{\(y = 0.1 (5\Phi((x - 2) / 0.3) + 1)\)} \\
    \hline
    0.5 & 0.100 & 0.938 & 0.947 \\
    \hline
    1.0 & 0.100 & 0.927 & 0.941 \\
    \hline
    1.5 & 0.124 & 0.826 & 0.830 \\
    \hline
    2.0 & 0.350 & 0.962 & 0.958 \\
    \hline
    2.5 & 0.576 & 0.835 & 0.816 \\
    \hline
    3.0 & 0.600 & 0.935 & 0.945 \\
    \hline
    3.5 & 0.600 & 0.957 & 0.953 \\
    \hline
    4.0 & 0.600 & 0.875 & 0.840 \\
    \hline
  \end{tabular}
\end{table}

\begin{table}[ht]
  \centering
  \caption{Results for Different Degrees of Freedom (\(n = 1000\))}
  \label{Table2}
  \begin{tabular}{|c|c|c|c|}
    \hline
    \textbf{x} & \textbf{Function} & \textbf{CP of df\_6} & \textbf{CP of df\_10} \\
    \hline
    \multicolumn{4}{|c|}{\(y = 0.25(x - 0.9)\)} \\
    \hline
    0.5 & -0.100 & 0.956 & 0.948 \\
    \hline
    1.0 & 0.025 & 0.953 & 0.957 \\
    \hline
    1.5 & 0.150 & 0.961 & 0.955 \\
    \hline
    2.0 & 0.275 & 0.967 & 0.955 \\
    \hline
    2.5 & 0.400 & 0.956 & 0.953 \\
    \hline
    3.0 & 0.525 & 0.949 & 0.961 \\
    \hline
    3.5 & 0.650 & 0.952 & 0.946 \\
    \hline
    4.0 & 0.775 & 0.952 & 0.937 \\
    \hline
    \multicolumn{4}{|c|}{\(y = 0.02(x^3+1.2)\)} \\
    \hline
    0.5 & 0.027 & 0.964 & 0.961 \\
    \hline
    1.0 & 0.044 & 0.950 & 0.954 \\
    \hline
    1.5 & 0.092 & 0.937 & 0.939 \\
    \hline
    2.0 & 0.184 & 0.939 & 0.949 \\
    \hline
    2.5 & 0.337 & 0.947 & 0.952 \\
    \hline
    3.0 & 0.564 & 0.952 & 0.946 \\
    \hline
    3.5 & 0.882 & 0.961 & 0.950 \\
    \hline
    4.0 & 1.30 & 0.963 & 0.936 \\
    \hline
    \multicolumn{4}{|c|}{\(y = 0.1 (5\Phi((x - 2) / 0.3) + 1)\)} \\
    \hline
    0.5 & 0.100 & 0.956 & 0.934 \\
    \hline
    1.0 & 0.100 & 0.936 & 0.912 \\
    \hline
    1.5 & 0.124 & 0.956 & 0.373 \\
    \hline
    2.0 & 0.350 & 0.967 & 0.951 \\
    \hline
    2.5 & 0.576 & 0.956 & 0.411 \\
    \hline
    3.0 & 0.600 & 0.954 & 0.905 \\
    \hline
    3.5 & 0.600 & 0.951 & 0.922 \\
    \hline
    4.0 & 0.600 & 0.946 & 0.837 \\
    \hline
  \end{tabular}
\end{table}

In table ~\ref{Table1}, \(x\) represents the certain time point we selected in the domain of the function, and \(f(x)\) is its true value calculated by the certain known function. CP represents coverage percentage of pointwise confidence interval we built by the bootstrap methods under certain degrees of freedom. We can see that in all those functions and degrees of freedom we have given, the bootstrap method gives pointwise confidence intervals coverage that are closed to the nominal level \(95\%\). And we can also see that while n gets larger, the result of \(n = 1000\), which can be seen in Table ~\ref{Table2} is much better than \(n = 100\) in each same level of degrees of freedom. While degrees of freedom is also an important variable inside the study and we can see that for the different degrees of freedom the result of CP is different either, but since it is not the main scope of this paper, we will not concern a lot about how to choose the best model to enhence the CP result.
